\thispagestyle{empty}

\begin{flushright}
    {\Huge\textit{``Now is better than never.''}}
    \linebreak
    \linebreak
    from ``The Zen of Python''
\end{flushright}

\clearpage

\thispagestyle{empty}

This thesis represents the end of a long journey started more than a decade
ago. Studying part-time has not been easy but it is definitively doable if one
has the right amount of perseverance, patience and determination: when I
enrolled to the master degree I was already working as software developer and I
have been working since then.

\medskip

It has been a bumpy ride, full of unexpected turns of events which slowed
down the progress of my studies but at the same time they contributed to my
personal and professional life and they brought me where I am now: 2 more
spoken languages, new home country, new culture, new experiences, plenty of
working experience and, the most important, working on AI/ML field, a
long-awaited move.

I have always seen the master course as way to learn new things (even from the
``boring'' topics), to improve myself and to keep my mind in constant training.
Although stressful, and sometimes hard, I would do it again.

\medskip

This has been possible thanks to the support of many people who in a way or
another have helped me to go through difficult times.

First of all to my family who has always believed and trusted me: their support
has been vital to get to the point where I am today.

My girlfriend Carmen who unconditionally has supported me during the last few
years for the final sprint of exams and thesis. Gracias mi amor!

Friends and people who, even with small words or actions, have given me the
strength to continue and to finish this journey.

\medskip

All the professors who, during all these years, have understood my working
abroad situation demonstrating flexibility and accommodating my needs.
Amongst them, a special thanks to professor Alfredo Milani, the supervisor of
this thesis.

\medskip

Last but not least, a huge thank you to my employer Arm Ltd. They have been
very supportive with my studies during all these years allowing me extra days
whenever I needed to sit exams and for allowing me to work part-time during my
final sprint.

Within Arm, I would like to thank my current line manager and supervisor of
this thesis, Anton Kachatkou: his support has been fundamental for the
execution of this work. Anton supported and helped me to get the right
approvals within Arm, IP reviewed every single word/code of this thesis, and
spent some of his time to discuss about this work.

I also wish to thank my team members, Mohamed Nour Abouelseoud, Ruomei Yan and
Johan Gras for reviewing my code and helping with experiments.

After all, it is a team effort and the excellent results of this thesis are
because of everyone's contribution.

\medskip

Thank you all.

\clearpage
